Using Model Checking to Analyze the System Behavior of the LHC Production Grid

### 2013-02-11 ElsevierFGCS_singleColumn.tex
Trace mining added to Related Work. Deductive verificaiton TBD.

To overcome the fundamental problem of state space explosion, several
variants to explicit model checking have gained popularity. Runtime verification~\cite{citeulike:10198274,Havelund:2008:RVC:1424465.1424470}
aims to increase confidence in the correctness, but does not make claims for absence of defects in a system.
The target program is instrumented manually, or automatically by scanning function calls and variable assignments, 
in order to capture and emit relevant events which are then monitored against a special component during execution.
This component resembles our monitoring process automaton used to cope with state space explosion during model checking 
of DIRAC's WMS system. While runtime verification is useful for partially-capturing incorrect behavior of the system implementation, 
code instrumentation is not always a feasible option, considering the increase of memory consumption. 
Similar to this lightweight form of verification is static program analysis~\cite{Schmidt98programanalysis}. Rather than code instrumentation and execution,
static program analysis aims to determine runtime properties of systems at compilation time.
A special analyzer targeted for the implementation language maintains information on the control flow and function call graphs
of the software. Approximation is used by such tools in practice, in order to improve scalability. The obvious advantage
of this approach is that it works directly on the implementation. However, most static analysis tools aim at discovering 
more general implementation deficiencies, such as uninitialized variables, null pointer references, or out of bound array indices. 
Both runtime verification and static analysis can be seen as additions to
traditional model checking, rather than alternatives. Some progress has been 
made in the direction of combining the approaches~\cite{Java_PathFinder,Linux_driver,Brat01combiningstatic}. For instance, Java PathFinder~\cite{Java_PathFinder}
relies on static analysis to reduce the state space before systematically exploring it by executing the abstracted program. In~\cite{Brat01combiningstatic}
static analysis and model checking also operate in a feedback loop. The drawback of static analyzers is their 
practical availability for languages other than C and Java. To the best of our knowledge, no such tool exists for Python.
However, our approach of using code slicing to abstract the implementation shares similarities.

TODO: trace mining

\section{Conclusions}
\label{sec:Section_5}
In this paper we have applied a formal methods approach for analyzing CERN's 
DIRAC grid system. 
We used two components as case studies: the Workload
Management and the Storage Management System, which are the driving force 
of DIRAC.
By creating an abstract model, simulating, visualizing, and model checking it with the mCRL2 toolset,
we were able to gain insight into the system behavior and detect critical
race conditions and livelocks, which  
were confirmed to occur in the real system. 
Despite the continuous DIRAC development for 10 years, these problems 
were notoriously difficult to discover and trace, considering the number of concurrent components involved. 
Testing of concurrent systems and reproducing error traces is often challenging because of the lack of 
control that the tester has over the execution of the concurrent processes. 
With the model at hand, replaying
the traces and localizing the problems became much more straightforward, overcoming many limitations of
traditional testing techniques for large-scale distributed systems.

>>>>

To overcome the problem of state space explosion, several
variants to explicit model checking have gained popularity. Runtime
verification~\cite{citeulike:10198274,Havelund:2008:RVC:1424465.1424470}
aims to increase confidence in the correctness, but does not make claims for
absence of defects in a system.
The target program is instrumented manually, or automatically by scanning
function calls and variable assignments, 
in order to capture and emit relevant events which are then monitored against a
special component during execution.
This component resembles our monitoring process automaton used to cope with
state space explosion during model checking 
of DIRAC's WMS system. While runtime verification is useful for
partially-capturing incorrect behavior of the system implementation, 
code instrumentation is not always a feasible option, considering the increase
of memory consumption.

Similar to this lightweight form of verification is static program
analysis~\cite{Schmidt98programanalysis}. Rather than code instrumentation and
execution, the objective of 
static program analysis is to determine runtime properties of systems at
compilation time.
A special analyzer targeted for the implementation language maintains
information on the control flow and function call graphs
of the software. Approximation is used by such tools in practice, in order to
improve scalability. The obvious advantage
of this approach is that it works directly on the implementation. However, most
static analysis tools aim at discovering 
more general implementation deficiencies, such as uninitialized variables, null
pointer references, or out of bound array indices. An additional drawback
is their 
practical availability for languages other than C and Java. To the best of our
knowledge, no such tool exists for Python.
Both runtime verification and static analysis can be seen as additions to
traditional model checking, rather than alternatives. Some progress has been 
made in the direction of combining the
approaches~\cite{Java_PathFinder,Linux_driver,Brat01combiningstatic}. For
instance, Java PathFinder~\cite{Java_PathFinder}
relies on static analysis to reduce the state space before systematically
exploring it by executing the abstracted program.
In~\cite{Brat01combiningstatic}
static analysis and model checking also operate in a feedback loop.

Instead of constructing a model of the system from scratch, when the only 
source of documentation is the code itself, a technique called process (or workflow) mining~\cite{Agrawal:1998:MPM:645338.650397}
can be used. It is a method of automatically distilling a process or software specification from a set of 
structured traces collected from real system executions.
The mined specifications are typically expressed in a form of Message Sequence Graphs or Petri-Nets, 
graphic formalism variants of the classical finite-state machines. Given that the 
analysis of program execution traces is still the most prevalent debugging technique
for distributed systems, and DIRAC's components (as well as the middleware) produce
immense amounts of execution logs anyhow, it seems like an attractive approach to constructing models.
There is a vast amount of work in this area~\cite{Aalst:1997:VWN:647744.733919,Lou:2010:MPW:1835804.1835883,Walkinshaw:2008:IFM:1642931.1642962,vanderAalst:2004:WMD:1018031.1018329,Ammons:2002:MS:503272.503275,Kumar:2011:SMC:1985793.1986002}.
A common assumption in process mining is that each trace is totally ordered, 
which for concurrent systems does not always hold. To cope with this, most approaches 
require that the thread IDs or process IDs are recorded in the traces as well.
Observing this as a realistic obstacle for obtaining an accurate model, Lou et al. in~\cite{Lou:2010:MPW:1835804.1835883}
propose an algorithm for constructing models from traces of concurrent systems by expressing events as partially ordered.
Their algorithm identifies dependency relationships between pairs of events in interleaved
traces. Another common assumption is a direct result of the underlying techniques used for constructing the models, namely data mining 
and machine learning. These techniques consider ``erroneous activities'' or activities that
were not logged for some reason, as noise, and disregard trace sequences with no statistical significance.
As a result, a design flaw can be mistaken for a noise, or not captured at all if a particular 
execution path has a low probability of occurrence. A notable exception in this direction is~\cite{vanderAalst:2005:PMV:2099016.2099032}, where the emphasis is not only on process discovery,
but also on verification: desired or undesired properties can be expressed with Linear Temporal Logic, 
and the checker verifies whether the observed behavior in the traces matches the property.

While relying solely on process mining as a technique for extracting a formal model may not be 
the way to go, inferring a ``fuzzy'' model from the frequent interaction traces, rather than manual code analysis
could be a helpful initial step. Nevertheless, the graphical models obtained must be 
translated into a process algebra formalism accepted by an actual toolset, if system correctness is to be established.
\section{Conclusions}
\label{sec:Section_5}
In this paper we have applied a formal methods approach for analyzing CERN's 
DIRAC grid system. 
We used two components as case studies: the Workload
Management and the Storage Management System, which are the driving force 
of DIRAC.
By creating an abstract model, simulating, visualizing, and model checking it with the mCRL2 toolset,
we were able to gain insight into the system behavior and detect critical
race conditions and livelocks, which  
were confirmed to occur in the real system. 
Despite the continuous DIRAC development for 10 years, these problems 
were notoriously difficult to discover and trace, considering the number of concurrent components involved. 
Testing of concurrent systems and reproducing error traces is often challenging because of the lack of 
control that the tester has over the execution of the concurrent processes. 
With the model at hand, replaying
the traces and localizing the problems became much more straightforward, overcoming many limitations of
traditional testing techniques for large-scale distributed systems.
